The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-10 rational.tex
reorg
algebraics

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Future Work}

The field of algebraic numbers consist of the rationals extended with
solutions to polynomials. In a limited sense {{langRevEE}} with
recursive types already include polynomials such as the definition of
binary tree {{@@x.(1+x*x)}} implies the equality {{t = 1 + t*t}} and
hence {{t^2-t+1=0}}. While this equation does indeed have an
irrational imaginary root, as we will see -- {{langRevEE}} does not
have the means of expressing meaningful computation that uses the
root.

Here is one way to view computations involving irrationals that might
shed some insight into how we should view recursive data types
also. For finite datatypes say of size {{n}} we can say that an
element specifies a choice of {{1/n}} and hence contains as much
information. For finite things, each value is computationally equal to
any other value. Any permutation is possible among the values. Hence
there is no metric possible on the values.  Recursive data types, such
as binary trees, have an infinite number of elements and hence the
argument that says that value specifies {{1/n}} units of information
is not longer valid. Instead to talk meaningfully about information
content, we should talk about a probability distribution on the values
of the type. However what shall we base this distribution on? Is there
a natural order beyond the unfolding order of the values of a type?

In this context, let us go back to the thought that recursive such as
{{nat=nat+1}} have no meaningful solutions, whereas trees do. But the
interesting solution of trees is precisely that they involve
irrational numbers. Our usual notation for irrationals such as
{{sqrt(2) }} can actually be viewed as a thunk representing the
infinite computation {{1.4142..}}. Irrationals in this sense have a
very different information theoretic interpretation from full real
numbers. With Chaitin's Omega numbers he postulates the existence of
real numbers that have an infinite amount of information -- ones for
example are the solution to the question about the halting properties
of every Turning machine.

Real numbers in that sense really are a series of bits of unknown
information theoretic interpretation -- a sort of {{top}}. Numbers
such as {{sqrt(2) }} are very different in the sense that even though
they contain an infinite sequence of bits, there is a finite program
that generates those bits and its only the cost of computing bits that
is infinite.  The consistency of arithmetic isn't affected by
extending the rational field by adding any specific computable
irrational. So it might be possible to design a model of computing
where we can indeed deal with them. Such a model will have to explicit
thunks for the delayed computation represented by trees and never
equate and cancel two infinite computations unless the represent
exactly the same infinite computation.

TODO.

Square root and imaginary types have also appeared in the literature: we
relegate the connections to Sec.~\ref{sec:related} and proceed with a simple
explanation. We have so far extended the set of types to be the rational
numbers. Now we will push this and extend the set of types to algebraic
numbers. In other words, we will allow datatypes defined by arbitrary
polynomials and allow the roots of such polynomials to be types.

Consider first an example in which we want to compute with the sides of a
rectangle whose area is 91 and whose length is longer than its width by 6
units. One can solve the quadratic equation to determine that the sides are 7
and 13 and proceed. This however prematurely forces us to globally solve the
constraint. Instead we can let the two sides of the rectangle be $x$ and
$x+6$ and use the following equation to capture the desired constraint:
\[
x^2 + 6x - 91 = 0
\]
The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
the type $0$. We can now proceed to compute with the unknown $x$, being
assured that in a closed program, our computation will eventually be
consistent with the solution of the algebraic equation.

Previously the most famous example of a similar nature is the puzzling
isomorphism that one can establish between seven binary trees and one.
A binary tree is defined by the datatype
\[
x = 1 + x * x 
\]
which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
algebraic manipulation we can reason as follows:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
Fiore poses the question of why such algebraic manipulation would make sense
type theoretically but states that even though some of the intermediate steps
make no sense, the final equivalence is valid and can be used to actually
construct an isomorphism between $x^7$ the type of seven binary trees 
and $x$ the type of binary trees.

Discussion of possible polynomials:
\begin{itemize}
\item $b=1+1$. Boring.
\item $b=b+1$. That introduces the natural numbers. No solution for this
  polynomial over the algebraic numbers. We must extend the numbers with
  $\omega$ to get a solution. We reject this in this paper and prefer to
  stick to algebraic numbers. The advantage is that all the isomorphisms are
  valid numerically. With the above type we could subtract $b$ from both
  sides to show that $0=1$ which is nonsense. We can however have infinite
  types as long as they have algebraic solutions.
\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
  boolean! If we had superpositions, we could then write a function that
  performs the square root of negation in the sense that applying it twice
  would be equivalent to boolean negation.
\end{itemize}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

-- WAVEFRONT START --

-- WAVEFRONT STOP --

The field of algebraic numbers consist of the rationals extended with
solutions to polynomials. In a limited sense {{langRevEE}} extended
with recursive types already include polynomials such as the
definition of binary tree {{@@x.(1+x*x)}} implies the equality
{{x=1+x*x}} and hence {{x^2-x+1=0}}.

%% Partial isomorphisms algebraics
%% do have computational

%% While this equation does indeed have an irrational imaginary root,
%% as we will see -- {{langRevEE}} does not have the means of
%% expressing meaningful computation that uses the root.

The famous example of isomorphism on algebraics is the embedding of
seven binary trees in one. In Figure 1 of Fiore's paper
\cite{Fiore:2004}, he demonstrates this isomorphism by the
corrsponding algebraic manipulation of the types.  The algebraic
manipulation can be viewed as a {{langRevEE}} combinator and it indeed
does constructively witness the isomorphism -- i.e. you can run it and
view one data structure encoded by the other. In this context,
consider the following algebraically valid proof of the same
isomorphism:

%% Previously the most famous example of a similar nature is the puzzling
%% isomorphism that one can establish between seven binary trees and one.
%% A binary tree is defined by the datatype
%% \[
%% x = 1 + x * x 
%% \]
%% which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
%% algebraic manipulation we can reason as follows:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
Fiore poses the question of why such algebraic manipulation would make
sense type theoretically, even though some of the intermediate steps
make no sense.  Stated in terms of {{langRevEE}} computations, this
manipulation has no meaningful computational content -- it is always
non-terminating. It suggests that some algebraic manipulations are
somehow more `constructive' than others or dually that {{langRevEE}}
with recursive types somehow lacks the ability to express all
computations that are otherwise algebraically meaningful.

-- WAVEFRONT START --

But the interesting solution of trees is precisely that they involve
irrational numbers. Our usual notation for irrationals such as
{{sqrt(2) }} can actually be viewed as a thunk representing the
infinite computation {{1.4142..}}.  Irrationals in this sense have a
very different information theoretic interpretation from full real
numbers.  With Chaitin's $\Omega$ numbers
\cite{Chaitin:1975:TPS:321892.321894} he postulates the existence of
real numbers that have an infinite amount of information -- ones for
example are the solution to the question about the halting properties
of every Turning machine.

That introduces the natural numbers. No solution for this polynomial
over the algebraic numbers. We must extend the numbers with $\omega$
to get a solution. We reject this in this paper and prefer to stick to
algebraic numbers. The advantage is that all the isomorphisms are
valid numerically. With the above type we could subtract $b$ from both
sides to show that $0=1$ which is nonsense. We can however have
infinite types as long as they have algebraic solutions.

-- WAVEFRONT STOP --

%% the final equivalence is valid and can be used to actually
%% construct an isomorphism between $x^7$ the type of seven binary
%% trees and $x$ the type of binary trees.

%% Here is one way to view computations involving irrationals that might
%% shed some insight into how we should view recursive data types
%% also. For finite datatypes say of size {{n}} we can say that an
%% element specifies a choice of {{1/n}} and hence contains as much
%% information. For finite things, each value is computationally equal to
%% any other value. Any permutation is possible among the values. Hence
%% there is no metric possible on the values.  Recursive data types, such
%% as binary trees, have an infinite number of elements and hence the
%% argument that says that value specifies {{1/n}} units of information
%% is not longer valid. Instead to talk meaningfully about information
%% content, we should talk about a probability distribution on the values
%% of the type. However what shall we base this distribution on? Is there
%% a natural order beyond the unfolding order of the values of a type?

%% In this context, let us go back to the thought that recursive such as
%% {{nat=nat+1}} have no meaningful solutions, whereas trees do.

%% Real numbers in that sense really are a series of bits of unknown
%% information theoretic interpretation -- a sort of {{top}}. Numbers
%% such as {{sqrt(2) }} are very different in the sense that even though
%% they contain an infinite sequence of bits, there is a finite program
%% that generates those bits and its only the cost of computing bits that
%% is infinite.  The consistency of arithmetic isn't affected by
%% extending the rational field by adding any specific computable
%% irrational. So it might be possible to design a model of computing
%% where we can indeed deal with them. Such a model will have to explicit
%% thunks for the delayed computation represented by trees and never
%% equate and cancel two infinite computations unless the represent
%% exactly the same infinite computation.

%% TODO.

%% Square root and imaginary types have also appeared in the literature: we
%% relegate the connections to Sec.~\ref{sec:related} and proceed with a simple
%% explanation. We have so far extended the set of types to be the rational
%% numbers. Now we will push this and extend the set of types to algebraic
%% numbers. In other words, we will allow datatypes defined by arbitrary
%% polynomials and allow the roots of such polynomials to be types.

%% Consider first an example in which we want to compute with the sides of a
%% rectangle whose area is 91 and whose length is longer than its width by 6
%% units. One can solve the quadratic equation to determine that the sides are 7
%% and 13 and proceed. This however prematurely forces us to globally solve the
%% constraint. Instead we can let the two sides of the rectangle be $x$ and
%% $x+6$ and use the following equation to capture the desired constraint:
%% \[
%% x^2 + 6x - 91 = 0
%% \]
%% The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
%% the type $0$. We can now proceed to compute with the unknown $x$, being
%% assured that in a closed program, our computation will eventually be
%% consistent with the solution of the algebraic equation.

@@
